{\huge Changes and trends in budburst and leaf flush across Europe and North America} \\A meta-analysis of local adaptation in spring phenology studies

### 2023-07-20 Docs/manuscript.Rnw
addressed comments

% recommended! Uncomment the below line and change the path for your computer!
\SweaveOpts{prefix.string=/Users/alina/Documents/git/localadaptclim/Docs/figure_ms/demoFig, eps=FALSE} 
%put your figures in one place! Also, note that here 'figures' is the folder and 'demoFig' is what each 
% figure produced will be titled plus its number or label (e.g., demoFig-nqpbetter.pdf')
% make your captioning look better
\usepackage[small]{caption}
\setlength{\captionmargin}{30pt}
\setlength{\abovecaptionskip}{0pt}
\setlength{\belowcaptionskip}{10pt}
% optional: muck with spacing
\topmargin -1.5cm        
\oddsidemargin 0.5cm   
\evensidemargin 0.5cm  % same as oddsidemargin but for left-hand pages
\textwidth 15.59cm
\textheight 21.94cm 
% \renewcommand{\baselinestretch}{1.5} % 1.5 lines between lines
\parindent 0pt		  % sets leading space for paragraphs
% optional: cute, fancy headers
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhead[LO]{March 2023}
\fancyhead[RO]{Manuscript}
% more optionals! %
% \usepackage[hyphens]{url} % this wraps my URL versus letting it spill across the page, a bad habit LaTeX has
%%% end preambling. %%%

@@

Based on these criteria we found 19 common gardens distributed throughout North America and Europe, with the majority of data concentrated in North America (Fig. \ref{figure:map_gardens} \& Supplement Table.\ref{supplement-table:all_studies}). From each common garden study we extracted phenological data on spring events (budburst and leaf flush) in DOY and, when present in the same paper, fall events (bud set, leaf senescence, growth cessation, and leaf abscission) by species and the geographic information of provenances and gardens. We used ImageJ (version 1.53k; \citealp{schneider_rasband_eliceiri_2012}) to extract values from figures whenever necessary. For studies that reported event dates relative to a reference date other than 1 January (e.g. \citealp{Rehfeldt1994}), we converted such dates to DOY using the `lubridate' package in R \citep{Grolemund11}. % emwJun4 -- you need to use ` to open single quotes in Latex. 
\begin{figure}[!h] 
    \centering
 \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/map_gardens.jpg}
    \caption{Distribution of common gardens (denoted as stars) and provenances (denoted as circles) included in this meta-analysis. The distribution was skewed toward North America (number of studies = 12). See Table.\ref{supplement-table:all_studies} in Supplements for information on selected studies.}  % emwJun4 -- add to caption how many studies in europe (7?)
    \label{figure:map_gardens}
\end{figure}
\newline
To understand how climatic differences, in addition to geographical differences, shape local adaptation in spring events we extracted several types of climate data using information about provenance latitude, longitude, and elevation from original publications. We estimated the mean annual temperature (MAT) for each provenance using the Climate Information Tool by Food and Agriculture Organization of the United Nations \citep{FAO2022}and ClimateWNA \citep{wang2016}.

To examine more explicitly climate near spring events, we used gridded daily temperature data for 2011 to 2020 for all European and North American provenances and gardens from E-OBS and the `daymetr' package in R respectively \citep{cornes2018,hufkens2018}. Using these data we estimated how much the daily temperatures overlapped between garden and provenance locations, which we call `climate overlap.' For this we used the 'overlap' package in R to calculate the percentage overlap of the daily temperature of each provenance and their corresponding gardens in spring months (March to May) from 2011 to 2020. Finally, using the daily temperature data, we also calculated Growing degree days (GDD), a commonly used heat accumulation measure to forecast phenological development in plants (Miller et al., 2001), based on the accumulation of mean daily temperatures ($T_{m}$)from 2011 to 2020 above a baseline of 0$^{\circ}$C, from January 1 until budburst and leaf flush with the following formula: GDD = $\sum$($T_{m}$ - $0^{\circ}$C for $T_{m}$ $\ge$ $0^{\circ}$C; 0 for $T_{m}$ $\le$ $0^{\circ}$C). % emwJun4 -- can you clarify in the text how you did GDD ... did you calculate it specific per tree or garden or ... ? And what about data before 2011? (I know we discussed, but I have forgotten))

\subsection{Analyses}
To estimate clines in spring and fall phenological events across species we used Bayesian hierarchical models. We regressed DOY of events against geographical and climate predictors with partial pooling (sometimes called `random effects') on the intercept and slope for each species within each garden. Because most tree species were present in only one common garden in our dataset, it was impossible to fit garden and species separately, thus we treat each species within a garden as a unique group. Using posterior estimates for each species within a garden, we estimated effects of continent (North America vs. Europe) and species type (angiosperm vs. gymnosperm). All models were fit in rstanarm package (version 2.21.3; \citealp{brilleman2018}) using default priors, with 4 chains and 1000 sampling iterations per chain for a total of 4000 samples. We checked for model fit by ... [ADD here]. We present estimates as mean $\pm$ 90\% uncertainty intervals given parenthetically, unless otherwise stated. 
%emw2Mar -- It would be good if you have tidy code you can post with the data so folks can fit the models themselves
% emwJun4 -- do you need help with ADD here ... if so, ping me in the old issue and we can work on it. It should be easy if you have cleared up all divergent transitions!
% We pooled the timing of budburst and leaf flush into a single category of ‘spring events’ and the timing of bud set, leaf senescence, growth cessation, and leaf abscission into ‘fall events.’ Such pooling is justified because of the shared pressures from natural selection that govern these events \citep{Gill15}.

@@

Overall, spring events were not related to provenance latitude or MAT, neither across continents (latitude: \Sexpr{format(round(mean(all_species_springpost_fitC_spring_lat),digits = 2), nsmall=2)} days/degree [\Sexpr{format(quantile(all_species_springpost_fitC_spring_lat, 0.05), digits = 1)} - \Sexpr{format(quantile(all_species_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT: \Sexpr{format(round(mean(all_species_springpost_fitC_spring_mat),digits = 2), nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(all_species_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(all_species_springpost_fitC_spring_mat, 0.95), digits = 2)}]) (Fig. \ref{figure:springfall_latmat}, Table. \ref{supplement-table:model_spring_lat} \& \ref{supplement-table:model_spring_mat} in Supplements), nor within North America (latitude: \Sexpr{format(round(mean(northamerican_springpost_fitC_spring_lat),digits = 2), nsmall=2)} days/degree [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.05), digits = 1)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT: \Sexpr{format(mean(northamerican_springpost_fitC_spring_mat),digits = 1)} days/$^{\circ}$C [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.95), digits = 2)}]) or Europe (latitude: \Sexpr{format(round(mean(european_springpost_fitC_spring_lat),digits = 2),nsmall=2)} days/degree [\Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT:\Sexpr{format(round(mean(european_springpost_fitC_spring_mat),digits = 2),nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.95), digits = 2)}]). Results were similar using other distance metrics in lieu of latitude (see Supplement Fig.\ref{supplement-figure:lat_distance} for results using the difference between provenance and garden latitude, and the spherical distance between provenance and garden).
% Spring events advanced slightly with provenance latitude in Europe spring events are slightly earlier where provenance MAT is lower (i.e., higher, more northern latitudes), though uncertainty intervals strongly overlapped zero. --- no longer the case

@@

In contrast, fall events (e.g., budset, leaf senescence, leaf abscission) advanced strongly with provenance latitude (\Sexpr{format(round(abs(mean(all_species_fallpost_fitC_fall_lat)),digits = 2), nsmall=2)} days/degree [\Sexpr{format(round(abs(quantile(all_species_fallpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)}-\Sexpr{format(round(abs(quantile(all_species_fallpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}]) and the decrease in MAT (\Sexpr{format(round(abs(mean(all_species_fallpost_fitC_fall_mat)),digits = 2), nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(all_species_fallpost_fitC_fall_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(all_species_fallpost_fitC_fall_mat, 0.95), digits = 2)}]), meaning fall events were earlier where provenance MAT was lower (Fig. \ref{figure:springfall_latmat}, Table.\ref{supplement-table:model_fall_lat} \& \ref{supplement-table:model_fall_mat} in Supplements ). This relationship, however, was observed mostly in North America where fall events advanced \Sexpr{format(round(abs(mean(northamericanpost_fitC_fall_lat)),digits = 2),nsmall=2)} (\Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree northward, or \Sexpr{format(round(mean(northamericanpost_fitC_fall_mat),digits = 2),nsmall=2)} days (\Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_mat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_mat, 0.95)), digits = 2),nsmall=2)}) per degree decline in MAT ($^{\circ}$C), whereas in Europe these relationships were weaker: advance of \Sexpr{format(round(abs(mean(europeanpost_fitC_fall_lat)),digits = 2),nsmall=2)} (\Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree northward, or \Sexpr{format(round(mean(europeanpost_fitC_fall_mat),digits = 2),nsmall=2)} days (\Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_mat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_mat, 0.95)), digits = 2),nsmall=2)}) per degree decline in MAT ($^{\circ}$C) (Fig. \ref{figure:continental_spptype_effect}A).

% Note to us, see also https://github.com/lizzieinvancouver/localadaptclim/issues/25
%7July2023: Add to this please! Add refs to figure and add the correct bibtex refs.
 \item Clines in fall phenology were stronger and more consistent whereas clines in spring phenology are weaker and somewhat varied in directionality. For fall events, only two field studies found no relationship (and ref Fig 2c-d; CITE Rosique-Esplugas 2021, Fraxinus excelsior from garden LETTER, which was in the UK and Petkova et al., 2017 Fagus sylvatica from garden LETTER, which was in Bulgaria, ), while another study was the only greenhouse experiment included (Rehfeldt 1994, also included Aitken and Bammals (2015)), which uniquely used the fall event of `the day by which seedling elongation had finished.'  Spring event clines were always weak, ... 
 %7July2023: Do any of the spring event clines for individual species x garden actually exclude 0 in their 90% interval? If not, add here:
 % (all species x garden clines included 0 in their 90\% intervals).
 % If so do not then add:
% ... were weak. Only XX and YY showed clines (give values with intervals) whereas the rest did not (all remaining species x garden clines included 0 in their 90\% intervals).

\begin{figure}[!h] 
    \centering
 \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/continental_spptype_effect.jpg}
    \caption{Posterior distributions for the effect of MAT across different continents and species types. Zero -- no effect -- is shown with a dashed line. Solid line and shading in posteriors represent mean and 50 percent interval. (A) Effect of MAT and continent on spring and fall event date (DOY). Fall events advanced strongly with decreasing MAT, particularly notably in North America. Effect of MAT and leaf type on spring and fall event date (DOY) weakly diverged. (B) Effect of MAT and leaf type on spring and fall event date (DOY). As MAT increased, spring events slightly advanced in angiosperms and delayed in gymnosperms. Fall events delayed in warmer locations for both species types.}
    \label{figure:continental_spptype_effect}
\end{figure}

% emwJun4 -- can you reference the figures in the supp that have the latitude results?
%7July2023 -- supp figures? Please add ref to them. 
Effects of provenance latitude on both spring and fall events were similar across angiosperms and gymnosperms.
Spring events slightly delayed, at a rate of \Sexpr{format(round(abs(mean(deciduous_springpost_fitC_spring_lat)),digits = 2), nsmall=2)} (\Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_lat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_lat, 0.95)), digits = 2),nsmall=2)}) days per degree north in angiosperms and \Sexpr{format(round(abs(mean(evergreens_springpost_fitC_spring_lat)),digits = 2), nsmall=2)} (\Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_lat, 0.05)), digits = 2),nsmall=2)}) days per degree north in gymnosperms. Fall events advanced \Sexpr{format(round(abs(mean(deciduouspost_fitC_fall_lat)),digits = 2), nsmall=2)} (\Sexpr{format(round(abs(quantile(deciduouspost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(deciduouspost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree north in angiosperms and \Sexpr{format(round(abs(mean(evergreenspost_fitC_fall_lat)),digits = 2), nsmall=2)} (\Sexpr{format(round(abs(quantile(evergreenspost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)}-\Sexpr{format(round(abs(quantile(evergreenspost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree north in gymnosperms.
However, effects of MAT on spring events weakly diverged (Fig.\ref{figure:continental_spptype_effect}B). Spring events advanced \Sexpr{format(round(abs(mean(deciduous_springpost_fitC_spring_mat)),digits = 2), nsmall=2)} (\Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_mat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_mat, 0.05)), digits = 2),nsmall=2)}) days/$^{\circ}$C as MAT increased in angiosperms and delayed \Sexpr{format(round(abs(mean(evergreens_springpost_fitC_spring_mat)),digits = 2), nsmall=2)} (\Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_mat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_mat, 0.95)), digits = 2),nsmall=2)}) days/$^{\circ}$C as MAT increased in gymnosperms. Fall events delayed in warmer locations for both species types, but slightly more so for gymnosperms (\Sexpr{format(round(mean(evergreenspost_fitC_fall_mat),digits = 2),nsmall=2)} days) than angiosperms (\Sexpr{format(round(mean(deciduouspost_fitC_fall_mat),digits = 2),nsmall=2)} days) (Fig. \ref{figure:continental_spptype_effect}B).

>>>>

% recommended! Uncomment the below line and change the path for your computer!
\SweaveOpts{prefix.string=/Users/alina/Documents/git/localadaptclim/Docs/figure_ms/demoFig, eps=FALSE} 
%put your figures in one place! Also, note that here 'figures' is the folder and 'demoFig' is what each 
% figure produced will be titled plus its number or label (e.g., demoFig-nqpbetter.pdf')
% make your captioning look better
\usepackage[small]{caption}
\setlength{\captionmargin}{30pt}
\setlength{\abovecaptionskip}{0pt}
\setlength{\belowcaptionskip}{10pt}
% optional: muck with spacing
\topmargin -1.5cm        
\oddsidemargin 0.5cm   
\evensidemargin 0.5cm  % same as oddsidemargin but for left-hand pages
\textwidth 15.59cm
\textheight 21.94cm 
% \renewcommand{\baselinestretch}{1.5} % 1.5 lines between lines
\parindent 0pt		  % sets leading space for paragraphs
% optional: cute, fancy headers
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhead[LO]{July 2023}
\fancyhead[RO]{Manuscript}
% more optionals! %
% \usepackage[hyphens]{url} % this wraps my URL versus letting it spill across the page, a bad habit LaTeX has
%%% end preambling. %%%

@@

Based on these criteria we found 19 common gardens distributed throughout North America and Europe, with the majority of data concentrated in North America (Fig. \ref{figure:map_gardens} \& Supplement Table.\ref{supplement-table:all_studies}). From each common garden study we extracted phenological data on spring events (budburst and leaf flush) in DOY and, when present in the same paper, fall events (bud set, leaf senescence, growth cessation, and leaf abscission) by species and the geographic information of provenances and gardens. We used ImageJ (version 1.53k; \citealp{schneider_rasband_eliceiri_2012}) to extract values from figures whenever necessary. For studies that reported event dates relative to a reference date other than 1 January (e.g. \citealp{Rehfeldt1994}), we converted such dates to DOY using the `lubridate' package in R \citep{Grolemund11}. % emwJun4 -- you need to use ` to open single quotes in Latex. 
\begin{figure}[!h] 
    \centering
 \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/map_gardens.jpg}
    \caption{Distribution of common gardens (denoted as stars) and provenances (denoted as circles) included in this meta-analysis. The distribution was skewed toward North America (12 North American studies versus 5 European studies). See Table.\ref{supplement-table:all_studies} in Supplements for information on selected studies.} 
    \label{figure:map_gardens}
\end{figure}
\newline
To understand how climatic differences, in addition to geographical differences, shape local adaptation in spring events we extracted several types of climate data using information about provenance latitude, longitude, and elevation from original publications. We estimated the mean annual temperature (MAT) for each provenance using the Climate Information Tool by Food and Agriculture Organization of the United Nations \citep{FAO2022}and ClimateWNA \citep{wang2016}.

To examine more explicitly climate near spring events, we used gridded daily temperature data for 2011 to 2020 for all European and North American provenances and gardens from E-OBS and the `daymetr' package in R respectively \citep{cornes2018,hufkens2018}. Using these data we estimated how much the daily temperatures overlapped between garden and provenance locations, which we call `climate overlap.' For this we used the 'overlap' package in R to calculate the percentage overlap of the daily temperature of each provenance and their corresponding gardens in spring months (March to May) from 2011 to 2020. Finally, using the daily temperature data, we also calculated growing degree days (GDD), a commonly used heat accumulation measure to forecast phenological development in plants \citep{miller01}, for each provenance and garden, based on the accumulation of mean daily temperatures ($T_{m}$)from 2011 to 2020 above a baseline of 0$^{\circ}$C, from January 1 until budburst and leaf flush with the following formula: GDD = $\sum$($T_{m}$ - $0^{\circ}$C for $T_{m}$ $\ge$ $0^{\circ}$C; 0 for $T_{m}$ $\le$ $0^{\circ}$C).

\subsection{Analyses}
To estimate clines in spring and fall phenological events across species we used Bayesian hierarchical models. We regressed DOY of events against geographical and climate predictors with partial pooling (sometimes called `random effects') on the intercept and slope for each species within each garden. Because most tree species were present in only one common garden in our dataset, it was impossible to fit garden and species separately, thus we treat each species within a garden as a unique group. Using posterior estimates for each species within a garden, we estimated effects of continent (North America vs. Europe) and species type (angiosperm vs. gymnosperm). All models were fit in rstanarm package (version 2.21.3; \citealp{brilleman2018}) using default priors, with 4 chains and 1000 sampling iterations per chain for a total of 4000 samples. We checked for model fit by ... [ADD here]. We present estimates as mean $\pm$ 90\% uncertainty intervals given parenthetically, unless otherwise stated.

% alinaJuly20 I need help with ADD here (ping'd you in the old issue. I have removed all divergent transitions already)

% We pooled the timing of budburst and leaf flush into a single category of ‘spring events’ and the timing of bud set, leaf senescence, growth cessation, and leaf abscission into ‘fall events.’ Such pooling is justified because of the shared pressures from natural selection that govern these events \citep{Gill15}.

@@

Overall, spring events were not related to provenance latitude or MAT, neither across continents (latitude: \Sexpr{format(round(mean(all_species_springpost_fitC_spring_lat),digits = 2), nsmall=2)} days/degree [\Sexpr{format(quantile(all_species_springpost_fitC_spring_lat, 0.05), digits = 1)} - \Sexpr{format(quantile(all_species_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT: \Sexpr{format(round(mean(all_species_springpost_fitC_spring_mat),digits = 2), nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(all_species_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(all_species_springpost_fitC_spring_mat, 0.95), digits = 2)}]) (Fig. \ref{figure:springfall_latmat}, Table. \ref{supplement-table:model_spring_lat} \& \ref{supplement-table:model_spring_mat} in Supplements), nor within North America (latitude: \Sexpr{format(round(mean(northamerican_springpost_fitC_spring_lat),digits = 2), nsmall=2)} days/degree [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.05), digits = 1)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT: \Sexpr{format(mean(northamerican_springpost_fitC_spring_mat),digits = 1)} days/$^{\circ}$C [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.95), digits = 2)}]) or Europe (latitude: \Sexpr{format(round(mean(european_springpost_fitC_spring_lat),digits = 2),nsmall=2)} days/degree [\Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT:\Sexpr{format(round(mean(european_springpost_fitC_spring_mat),digits = 2),nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.95), digits = 2)}]) (Fig.\ref{figure:continental_spptype_lat}A \& \ref{figure:continental_spptype_effect}A). Results were similar using other distance metrics in lieu of latitude (see Supplement Fig.\ref{supplement-figure:lat_distance} for results using the difference between provenance and garden latitude, and the spherical distance between provenance and garden).
% Spring events advanced slightly with provenance latitude in Europe spring events are slightly earlier where provenance MAT is lower (i.e., higher, more northern latitudes), though uncertainty intervals strongly overlapped zero. --- no longer the case

@@

In contrast, fall events (e.g., budset, leaf senescence, leaf abscission) advanced strongly with provenance latitude (\Sexpr{format(round(abs(mean(all_species_fallpost_fitC_fall_lat)),digits = 2), nsmall=2)} days/degree [\Sexpr{format(round(abs(quantile(all_species_fallpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)}-\Sexpr{format(round(abs(quantile(all_species_fallpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}]) and the decrease in MAT (\Sexpr{format(round(abs(mean(all_species_fallpost_fitC_fall_mat)),digits = 2), nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(all_species_fallpost_fitC_fall_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(all_species_fallpost_fitC_fall_mat, 0.95), digits = 2)}]), meaning fall events were earlier where provenance MAT was lower (Fig. \ref{figure:springfall_latmat}, Table.\ref{supplement-table:model_fall_lat} \& \ref{supplement-table:model_fall_mat} in Supplements ). This relationship, however, was observed mostly in North America where fall events advanced \Sexpr{format(round(abs(mean(northamericanpost_fitC_fall_lat)),digits = 2),nsmall=2)} (\Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree northward, or \Sexpr{format(round(mean(northamericanpost_fitC_fall_mat),digits = 2),nsmall=2)} days (\Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_mat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_mat, 0.95)), digits = 2),nsmall=2)}) per degree decline in MAT ($^{\circ}$C), whereas in Europe such relationship was weaker: advance of \Sexpr{format(round(abs(mean(europeanpost_fitC_fall_lat)),digits = 2),nsmall=2)} (\Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree northward, or \Sexpr{format(round(mean(europeanpost_fitC_fall_mat),digits = 2),nsmall=2)} days (\Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_mat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_mat, 0.95)), digits = 2),nsmall=2)}) per degree decline in MAT ($^{\circ}$C) (Fig. \ref{figure:continental_spptype_effect}A).

% Note to us, see also https://github.com/lizzieinvancouver/localadaptclim/issues/25

Clines in fall phenology were stronger and more consistent whereas clines in spring phenology are weaker and somewhat varied in directionality. For fall events, only two field studies found no relationship (Fig. \ref{figure:springfall_latmat}): \emph{Fraxinus excelsior} from Garden Q* in the UK \citep{rosique22} and \emph{Fagus sylvatica} from Garden R* in Bulgaria \citep{petkova17}. Another study that found no relationship was the only greenhouse experiment included (\emph{Picea engelmannii} from Garden B in the USA, also included in \citealp{AitkenBemmels16}), which uniquely used the fall event of `the day by which seedling elongation had finished' \citep{rehfeldt94}. Spring event clines were always weak: all species x garden clines included 0 in their 90\% intervals.

\begin{figure}[!h] 
    \centering
 \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/continental_spptype_lat.jpg}
    \caption{Posterior distributions for the effect of provenance latitude across different continents and species types. Zero -- no effect -- is shown with a dashed line. Solid line and shading in posteriors represent mean and 50 percent interval. (A) Effect of latitude and continent on spring and fall event date (DOY). Fall events advanced strongly per degree northward, particularly notably in North America. (B) Effect of latitude and leaf type on spring and fall event date (DOY). As we move north, spring events slightly advanced in gymnosperms and delayed in angiosperms. Fall events advanced per degree northward for both species types.}
    \label{figure:continental_spptype_lat}
\end{figure}

\begin{figure}[!h] 
    \centering
 \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/continental_spptype_effect.jpg}
    \caption{Posterior distributions for the effect of MAT across different continents and species types. Zero -- no effect -- is shown with a dashed line. Solid line and shading in posteriors represent mean and 50 percent interval. (A) Effect of MAT and continent on spring and fall event date (DOY). Fall events advanced strongly with decreasing MAT, particularly notably in North America. (B) Effect of MAT and leaf type on spring and fall event date (DOY). As MAT increased, spring events slightly advanced in angiosperms and delayed in gymnosperms. Fall events delayed in warmer locations for both species types.}
    \label{figure:continental_spptype_effect}
\end{figure}

Effects of provenance latitude on both spring and fall events were similar across angiosperms and gymnosperms (Fig.\ref{figure:continental_spptype_lat}B). Spring events weakly diverged, delaying at a rate of \Sexpr{format(round(abs(mean(deciduous_springpost_fitC_spring_lat)),digits = 2), nsmall=2)} (\Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_lat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_lat, 0.95)), digits = 2),nsmall=2)}) days per degree north in angiosperms and advancing \Sexpr{format(round(abs(mean(evergreens_springpost_fitC_spring_lat)),digits = 2), nsmall=2)} (\Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_lat, 0.05)), digits = 2),nsmall=2)}) days per degree north in gymnosperms. Fall events advanced \Sexpr{format(round(abs(mean(deciduouspost_fitC_fall_lat)),digits = 2), nsmall=2)} (\Sexpr{format(round(abs(quantile(deciduouspost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(deciduouspost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree north in angiosperms and \Sexpr{format(round(abs(mean(evergreenspost_fitC_fall_lat)),digits = 2), nsmall=2)} (\Sexpr{format(round(abs(quantile(evergreenspost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)}-\Sexpr{format(round(abs(quantile(evergreenspost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree north in gymnosperms.
Effects of MAT on spring events also weakly diverged (Fig.\ref{figure:continental_spptype_effect}B). Spring events advanced \Sexpr{format(round(abs(mean(deciduous_springpost_fitC_spring_mat)),digits = 2), nsmall=2)} (\Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_mat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_mat, 0.05)), digits = 2),nsmall=2)}) days/$^{\circ}$C as MAT increased in angiosperms and delayed \Sexpr{format(round(abs(mean(evergreens_springpost_fitC_spring_mat)),digits = 2), nsmall=2)} (\Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_mat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_mat, 0.95)), digits = 2),nsmall=2)}) days/$^{\circ}$C as MAT increased in gymnosperms. Fall events delayed in warmer locations for both species types, but slightly more so for gymnosperms (\Sexpr{format(round(mean(evergreenspost_fitC_fall_mat),digits = 2),nsmall=2)} days) than angiosperms (\Sexpr{format(round(mean(deciduouspost_fitC_fall_mat),digits = 2),nsmall=2)} days) (Fig. \ref{figure:continental_spptype_effect}B).

While we expected that coarse metrics, such as latitude and MAT, would generally represent how similar the climates are between the provenances and gardens, we also estimated climate overlap in months much closer to the events to further test how much climate similarity between provenances and gardens predicts local adaptation. For spring events, we considered overlap across March to May. However, results were not qualitatively different than using MAT (See Fig. \ref{supplement-figure:overlap_scatterplot} in Supplements). We observed very weak effects of climate overlap on spring events (\Sexpr{format(round(abs(mean(all_species_springpost_fitC_spring_overlap)),digits = 2), nsmall=2)} [\Sexpr{format(round(abs(quantile(all_species_springpost_fitC_spring_overlap, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(all_species_springpost_fitC_spring_overlap, 0.95)), digits = 2),nsmall=2)}] days per one per cent increase in climate overlap), nearly identical across angiosperms (\Sexpr{format(round(abs(mean(deciduous_springpost_fitC_spring_overlap)),digits = 2), nsmall=2)} [\Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_overlap, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_overlap, 0.05)), digits = 2),nsmall=2)}]) and gymnosperms (\Sexpr{format(round(abs(mean(evergreens_springpost_fitC_spring_overlap)),digits = 2), nsmall=2)} [\Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_overlap, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_overlap, 0.95)), digits = 2),nsmall=2)}]). Fall events advanced as climate overlap declined, but slightly more strongly for gymnosperms (advancing \Sexpr{format(round(abs(mean(evergreenspost_fitC_fall_overlap)),digits = 2), nsmall=2)} [\Sexpr{format(round(abs(quantile(evergreenspost_fitC_fall_overlap, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(evergreenspost_fitC_fall_overlap, 0.95)), digits = 2),nsmall=2)}] days per one per cent decline in climate overlap) (Fig.\ref{supplement-figure:overlap_posterior} in Supplements).
